Developing a Bubble Chamber Particle Discriminator Using Semi-Supervised Learning

### 2018-10-19 documentation/paper.tex
Scrapped a lot of stuff in the report

$WallCut(D)$ was applied to the low-resolution Fourier transforms during early experiments (since they were used for AP) to minimize the number of variables between initial neural networks and a proven discriminator.

This time, position corrections were optionally applied to the input data, as they were the input to Acoustic Parameter. The resulting configurations are furthermore referred to as $FourierMLP(\beta_{N})$, $FourierMLP(PosCor(\beta_{N}, \chi))$ and $FourierMLPWithPos(\beta_{N}, \chi)$. Several different resolutions $N$ of $\beta_{N}$ were used. While a variety of network architectures were tested, most had a small number of layers (on the order of three), applied dropout and L2 regularization, and used batch normalization once again.

@@

It became evident very quickly that, without $WallCut(D)$, a high resolution was not required to get a high accuracy relative to the ground truth data. $FourierMLP(\beta_{8})$ managed an excellent 98\% validation accuracy, and also produced a mean class-wise standard deviation of 0.29. Its very decisive prediction distribution can be observed in Figure \ref{banded_no_pos_input_hist}. Increasing the number of bands to $FourierMLPWithPos(beta_{20}, \chi)$, did not improve the results; accuracy only reached 95\%, with an increased mean standard deviation of 0.42. Using 40 bands reduced the accuracy once again, to 91\%.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{banded_no_pos_input_hist}
    \caption{\label{banded_no_pos_input_hist} Prediction distribution of the best $FourierMLP(\beta_{8})$ discriminator, with $WallCut(D)$.}
\end{figure}

Very interestingly, $FourierMLP(\beta_{8})$ performed better than either $FourierMLP(PosCor(\beta_{8}), \chi)$ or \\ $FourierMLPWithPos(\beta{8}, \chi)$, which produced accuracy values of 96\% and 97\%, respectively.

Unlike for the convolutional neural network trained on $\omega$, position input did not improve performance at all. In fact, $FourierMLP(\beta_{8})$, performed better than $FourierMLPWithPos(\beta_{8}, \chi)$, which reached only 95\% validation accuracy. Nevertheless, they both performed better than any of the network architectures trained on $\omega$.

The higher resolution demonstrated significant power when $WallCut(D)$ was not applied, and only \\ $QualCut(MultiCut(D))$ was used. $FourierMLPWithPos(\beta_{20}, \chi)$ achieved only 92\% accuracy (2 percentage points lower than the same configuration without $WallCut(D)$). However, training on $\beta_{50,001}$, which does not incorporate any banding, improved this figure to a maximum of 97\%. However, that epoch had a relatively high class-wise standard deviation, at 0.47, as seen in Figure \ref{high_freq_no_wall_hist}.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{high_freq_no_wall_hist}
    \caption{\label{high_freq_no_wall_hist} Prediction distribution of the best $FourierMLPWithPos(\beta_{50,001}, \chi)$ discriminator, without $WallCut(D)$.}
\end{figure}

@@

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Neural networks trained on Fourier transforms with $WallCut(D)$.} \label{fft_wall_cuts}
        \begin{tabular}{|l|l|l|l|l|l|l|l|}
            \hline
            Configuration & L2 $\lambda$ & Dropout & Layers & Max Acc & Prec & Recall & CWSD \\
            \hline
            $FourierMLP(PosCor(\beta_{8}))$ & 0 & 0.5 & 2 & 96\% & 98\% & 97\% & 0.42 \\
            % ../experimental_data/validation_results/2018_07_16_banded_frequency_validation_passing_weighted_3_times.json
            \hline
            $FourierMLP(\beta_{8})$ & 0 & 0.5 & 2 & 98\% & 100\% & 98\% & 0.29 \\
            % ../experimental_data/validation_results/2018_07_17_banded_frequency_not_position_corrected/banded_low_resolution_time1531831056_epoch249.json
            \hline
            $FourierMLPWithPos(\beta_{8}, \chi)$ & 0 & 0.5 & 3 & 95\% & 100\% & 92\% & 0.44 \\
            % ../experimental_data/validation_results/2018_07_17_banded_frequency_not_position_corrected_with_position_input/banded_low_resolution_time1531837503_epoch18.json
            \hline
            $FourierMLPWithPos(\beta_{8}, \chi)$ & 0 & 0.25 & 3 & 97\% & 98\% & 98\% & 0.43 \\
            % ../experimental_data/validation_results/2018_07_18_banded_frequency_25\%_dropout/banded_low_resolution_time1531922044_epoch154.json
            \hline
            $FourierMLPWithPos(\beta_{20}, \chi)$ & 0.003 & 0.25 & 3 & 93\% & 99\% & 92\% & 0.54 \\
            % ../experimental_data/validation_results/2018_07_20_20_bands/high_resolution_frequency_time1532105605_epoch90.json
            \hline
            $FourierMLPWithPos(\beta_{20}, \chi)$ & 0 & 0 & 3 & 94\% & 99\% & 93\% & 0.52 \\
            % ../experimental_data/validation_results/2018_07_20_20_bands_no_regularization/high_resolution_frequency_time1532113464_epoch172.json
            \hline
            $FourierMLPWithPos(\beta_{40}, \chi)$ & 0 & 0 & 3 & 91\% & 98\% & 90\% & 0.57 \\
            % ../experimental_data/validation_results/2018_07_20_40_bands/high_resolution_frequency_time1532107386_epoch101.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.0003 & 0 & 4 & 91\% & 94\% & 93\% & 0.53 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0_l2_lambda0.0003_hidden_layers3_time1533922958_epoch133.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.001 & 0.25 & 4 & 91\% & 96\% & 89\% & 0.56 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0.25_l2_lambda0.001_hidden_layers3_time1533924257_epoch115.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.001 & 0.5 & 3 & 89\% & 91\% & 91\% & 0.64 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0.5_l2_lambda0.001_hidden_layers2_time1533925299_epoch169.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.01 & 0.5 & 4 & 89\% & 90\% & 93\% & 0.65 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0.5_l2_lambda0.01_hidden_layers3_time1533925799_epoch91.json
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

@@

Without wall cuts, the full-resolution Fourier transform produces much more effective discriminators. Here, L2 regularization is beneficial, possibly because there is more noise that can be overfit on.

Unlike with the raw waveform, recall varies relatively little when wall cuts are applied.

@@

It is an open question whether or not there is any information in the image data $\iota$ that could be used to distinguish between particle classes. Relative to the extremely short period of time in which the bubble forms (on the scale of nanoseconds), the framerate of the camera (340Hz) is extremely slow. While the very early stages of bubble formation (when the sound is produced) are known to differ depending on whether the bubble was created by a nuclear recoil or an alpha particle, it is unknown whether any visually apparent differences persist when the bubble is visible.

@@

\begin{figure}[h]
    \centering
    \includegraphics[width=\textwidth]{image_hist}
    \caption{\label{image_hist} Prediction distribution of the best $ImageConv(\iota)$ discriminator.}
\end{figure}

@@

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Summary of supervised learning configurations, trained with $WallCut(D)$.}
        \begin{tabular}{|l|l|l|l|l|}
            \hline
            Configuration & Max Accuracy & Precision & Recall & CWSD \\
            \hline
            $DeepConvWithPos(\omega, \chi)$ & 96\% & 99\% & 96\% & 0.41 \\
            \hline
            $FourierMLP(PosCor(\beta_{8}))$ & 96\% & 98\% & 97\% & 0.42 \\
            \hline
            $FourierMLP(\beta_{8})$ & 98\% & 100\% & 98\% & 0.29 \\
            \hline
            $FourierMLPWithPos(\beta_{8}, \chi)$ & 97\% & 98\% & 98\% & 0.43 \\
            \hline
            $FourierMLPWithPos(\beta_{20}, \chi)$ & 94\% & 99\% & 93\% & 0.52 \\
            \hline
            $FourierMLPWithPos(\beta_{40}, \chi)$ & 91\% & 98\% & 90\% & 0.57 \\
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 91\% & 94\% & 93\% & 0.53 \\
            \hline
            Original PICO-60 Neural Network & 83\% & 83\% & 95\% & 0.83 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Summary of supervised learning configurations, trained without $WallCut(D)$.}
        \begin{tabular}{|l|l|l|l|l|}
            \hline
            Configuration & Max Accuracy & Precision & Recall & CWSD \\
            \hline
            $FourierMLPWithPos(\beta_{20}, \chi)$ & 92\% & 93\% & 77\% & 0.63 \\
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 97\% & 93\% & 98\% & 0.47 \\
            \hline
            $DeepConv(\omega)$ & 77\% & 76\% & 94\% & 0.86 \\
            % ../experimental_data/validation_results/2018_07_16_waveform_with_new_cuts/waveform_time1531762243_epoch14.json
            \hline
            $DeepConvWithPos(\omega, \chi)$ & 91\% & 95\% & 78\% & 0.61 \\
            \hline
            $ImageConv(\iota)$ & 85\% & 40\% & 38\% & 0.95 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

There are 2 key conclusions to be drawn from the above tables:

\begin{enumerate}
    \item With $WallCut(D)$, the low-resolution Fourier transform $\beta_{8}$ is by far the most effective input format. Higher resolutions seem to make performance worse, likely due to overfitting. It is significantly better than the original PICO-60 neural network, and is more decisive (as measured by the class-wise standard deviation)
    \item Without $WallCut(D)$, full-resolution $\beta_{50,001}$ is the most effective, nearing the best performance with wall cuts at separating run types. All models are notably less decisive, given that the lowest standard deviation without wall cuts is 0.47, compared to 0.29 with wall cuts.
\end{enumerate}

Based on these results, experimentation with semi-supervised learning proceeded using \\ $FourierMLP(PosCor(\beta_{8}))$, applying a 3-layer perceptron architecture, and using $WallCut(D)$. There were three main reasons for this:

\begin{enumerate}
    \item This was a highly effective configuration with wall cuts, producing 97\% accuracy and a class-wise standard deviation dramatically better than the original PICO-60 neural network on that data set.
    \item The use of the exact same input format as AP makes the algorithms more directly comparable.
\end{enumerate}

@@

Since the intent of these techniques is to separate particle types (the same as AP), performance evaluations (accuracy and class-wise standard deviation) use AP as a baseline. Note that $WallCut(D)$ was once again applied for semi-supervised learning, because $FourierMLP(PosCor(\beta_{8}))$ was used for a neural network configuration, and because AP requires application of $WallCut(D)$ to be meaningfully comparable to any neural network.

To ensure that accuracy statistics are as consistent and reliable as possible, every configuration was trained three times, with a different randomly selected validation set each time. Accuracy, precision, recall, and class-wise standard deviation statistics are the mean over those 3 tests.

>>>>

This time, position corrections were optionally applied to the input data, as they were the input to Acoustic Parameter. The resulting configurations are furthermore referred to as $FourierMLP(\beta_{N})$ and \\ $FourierMLP(PosCor(\beta_{N}, \chi))$. Several different resolutions $N$ of $\beta_{N}$ were used. While a variety of network architectures were tested, most had a small number of layers (on the order of three), applied dropout and L2 regularization, and used batch normalization once again.

@@

It became evident very quickly that a high resolution was not required to get a high accuracy relative to the ground truth data. $FourierMLP(\beta_{8})$ managed an excellent 98\% validation accuracy, and also produced a mean class-wise standard deviation of 0.29. Its very decisive prediction distribution can be observed in Figure \ref{banded_no_pos_input_hist}. Increasing the number of bands to $FourierMLPWithPos(beta_{20}, \chi)$, did not improve the results; accuracy only reached 95\%, with an increased mean standard deviation of 0.42. Using 40 bands reduced the accuracy once again, to 91\%.

Very interestingly, $FourierMLP(\beta_{8})$ performed better than $FourierMLP(PosCor(\beta_{8}), \chi))$, which produced an accuracy value of 96\%. Both of these results were better than any of the network architectures trained on $\omega$.

\begin{figure}[H]
    \centering
    \includegraphics[width=\textwidth]{banded_no_pos_input_hist}
    \caption{\label{banded_no_pos_input_hist} Prediction distribution of the best $FourierMLP(\beta_{8})$ discriminator.}
\end{figure}

@@

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Neural networks trained on Fourier transforms.} \label{fft_wall_cuts}
        \begin{tabular}{|l|l|l|l|l|l|l|l|}
            \hline
            Configuration & L2 $\lambda$ & Dropout & Layers & Max Acc & Prec & Recall & CWSD \\
            \hline
            $FourierMLP(PosCor(\beta_{8}))$ & 0 & 0.5 & 2 & 96\% & 98\% & 97\% & 0.42 \\
            % ../experimental_data/validation_results/2018_07_16_banded_frequency_validation_passing_weighted_3_times.json
            \hline
            $FourierMLP(\beta_{8})$ & 0 & 0.5 & 2 & 98\% & 100\% & 98\% & 0.29 \\
            % ../experimental_data/validation_results/2018_07_17_banded_frequency_not_position_corrected/banded_low_resolution_time1531831056_epoch249.json
            \hline
            $FourierMLPWithPos(\beta_{8}, \chi)$ & 0 & 0.5 & 3 & 95\% & 100\% & 92\% & 0.44 \\
            % ../experimental_data/validation_results/2018_07_17_banded_frequency_not_position_corrected_with_position_input/banded_low_resolution_time1531837503_epoch18.json
            \hline
            $FourierMLPWithPos(\beta_{8}, \chi)$ & 0 & 0.25 & 3 & 97\% & 98\% & 98\% & 0.43 \\
            % ../experimental_data/validation_results/2018_07_18_banded_frequency_25\%_dropout/banded_low_resolution_time1531922044_epoch154.json
            \hline
            $FourierMLPWithPos(\beta_{20}, \chi)$ & 0.003 & 0.25 & 3 & 93\% & 99\% & 92\% & 0.54 \\
            % ../experimental_data/validation_results/2018_07_20_20_bands/high_resolution_frequency_time1532105605_epoch90.json
            \hline
            $FourierMLPWithPos(\beta_{20}, \chi)$ & 0 & 0 & 3 & 94\% & 99\% & 93\% & 0.52 \\
            % ../experimental_data/validation_results/2018_07_20_20_bands_no_regularization/high_resolution_frequency_time1532113464_epoch172.json
            \hline
            $FourierMLPWithPos(\beta_{40}, \chi)$ & 0 & 0 & 3 & 91\% & 98\% & 90\% & 0.57 \\
            % ../experimental_data/validation_results/2018_07_20_40_bands/high_resolution_frequency_time1532107386_epoch101.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.0003 & 0 & 4 & 91\% & 94\% & 93\% & 0.53 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0_l2_lambda0.0003_hidden_layers3_time1533922958_epoch133.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.001 & 0.25 & 4 & 91\% & 96\% & 89\% & 0.56 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0.25_l2_lambda0.001_hidden_layers3_time1533924257_epoch115.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.001 & 0.5 & 3 & 89\% & 91\% & 91\% & 0.64 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0.5_l2_lambda0.001_hidden_layers2_time1533925299_epoch169.json
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 0.01 & 0.5 & 4 & 89\% & 90\% & 93\% & 0.65 \\
            % ../experimental_data/validation_results/2018_08_10_high_resolution_grid_search_with_wall_cuts/high_res_grid_search_dropout0.5_l2_lambda0.01_hidden_layers3_time1533925799_epoch91.json
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

@@

@@

It is an open question whether or not there is any information in the image data $\iota$ that could be used to distinguish between particle classes. Relative to the extremely short period of time in which the bubble forms (on the scale of nanoseconds), the framerate of the camera (340Hz) is extremely slow. While the very early stages of bubble formation (when the sound is produced) are known to differ depending on whether the bubble was created by a nuclear recoil or an alpha particle, it is not experimentally known whether any visually apparent differences persist when the bubble is visible.

@@

\begin{figure}[h]
    \centering
    % includegraphics[width=\textwidth]{image_hist}
    \caption{\label{image_hist} Prediction distribution of the best $ImageConv(\iota)$ discriminator.}
\end{figure}

@@

\begin{minipage}{\textwidth}
    \begin{center}
        \captionof{table}{Summary of supervised learning configurations.}
        \begin{tabular}{|l|l|l|l|l|}
            \hline
            Configuration & Max Accuracy & Precision & Recall & CWSD \\
            \hline
            $DeepConvWithPos(\omega, \chi)$ & 96\% & 99\% & 96\% & 0.41 \\
            \hline
            $FourierMLP(PosCor(\beta_{8}))$ & 96\% & 98\% & 97\% & 0.42 \\
            \hline
            $FourierMLP(\beta_{8})$ & 98\% & 100\% & 98\% & 0.29 \\
            \hline
            $FourierMLPWithPos(\beta_{20}, \chi)$ & 94\% & 99\% & 93\% & 0.52 \\
            \hline
            $FourierMLPWithPos(\beta_{40}, \chi)$ & 91\% & 98\% & 90\% & 0.57 \\
            \hline
            $FourierMLPWithPos(\beta_{50,001}, \chi)$ & 91\% & 94\% & 93\% & 0.53 \\
            \hline
            Original PICO-60 Neural Network & 83\% & 83\% & 95\% & 0.83 \\
            \hline
        \end{tabular}
    \end{center}
\end{minipage}

The key conclusion to be drawn from the above tables is that the low-resolution Fourier transform $\beta_{8}$ is by far the most effective input format. Higher resolutions seem to make performance worse, likely due to overfitting. It is significantly better than the original PICO-60 neural network, and is more decisive (as measured by the class-wise standard deviation).

Based on this result, experimentation with semi-supervised learning proceeded using \\ $FourierMLP(PosCor(\beta_{8}, \chi))$, applying a three-layer perceptron architecture. There were two main reasons for this:

\begin{enumerate}
    \item This was a highly effective configuration, producing 97\% accuracy and a class-wise standard deviation dramatically better than the original PICO-60 neural network on that data set.
    \item The use of the exact same input format as AP makes the algorithms more directly comparable.
\end{enumerate}

@@

Since the intent of these techniques is to separate particle types (the same as AP), performance evaluations (accuracy and class-wise standard deviation) use AP as a baseline. To ensure that accuracy statistics are as consistent and reliable as possible, every configuration was trained three times, with a different randomly selected validation set each time. Accuracy, precision, recall, and class-wise standard deviation statistics are the mean over those three tests.
